---
title: "Final Project: Part One R Graphics"
author: "Adam Bruce"
date: "02/02/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)
# feel free to change fig.height or fig.width here or in specific chunk headings. 
```


```{r}
# load packages - add any you would like
library(tidyverse)
library(lme4)
library(lmerTest)
library(gridExtra)
library(knitr)
library(dplyr)
library(ggplot2)
library(knitr)
library(gridExtra)
```

## Abstract 

A 150 word abstract.


\newpage

### Background and Significance   

Type background and significance section here. 

### Data  

Below are my five initial exploratory graphics and summaries of the Kentucky Math Score Dataset.  

Note KMS is Kentucky Math Scores


```{r}

# Note KMS is Kentucky Math Scores

KMS <- read_csv("https://raw.githubusercontent.com/proback/BeyondMLR/master/data/kentucky.csv")

head(KMS)


KMS <- na.omit(KMS)

```

### 1 Boxplot for Male vs Female Scores for ALL 48,058 students from ALL schools

```{r, fig.width = 10, fig.height = 6}

#1 Boxplot for Male vs Female Scores for ALL 48,058 students from ALL schools


KMS$female <- as.character(KMS$female)

ggplot(data = KMS, aes(x = female, y = mathn, color = female, fill = female))+
  geom_boxplot(size = 1.5)+
  scale_fill_manual(values=c("dodgerblue", "gold2"))+
  scale_color_manual(values=c("chartreuse", "darkorchid1"))+
  xlab("Sex")+ 
  ylab("California Math Assessment Score")+ 
  labs(title = "Comparison of California Math Assessment Scores for Female (1) vs Male (0) Kentucky 8th Grade Students")

  

```

### 2 Summary Table for Average Scores for the 132 Districts


The HEAD of the Kable created is displayed here (it would be too long if I displayed all 132 districts)

```{r}

#2 Summary Table for Average Scores for the 132 Districts

DISTRICT.KMS <- KMS %>% group_by(dis_id) %>% summarize(MEAN.MATH = mean(mathn), MEAN.SIZE = mean(sch_size), MEAN.SES = mean(sch_ses), MEAN.DLS = mean(dis_size), MEAN.DSES = mean(dis_ses))

# The HEAD of the Kable created is displayed here (it would be too long if I displayed all 132 districts)

kable(head(DISTRICT.KMS))


```


### 3 Scatter Plots of Non-Averages VS. Averages, District Size & Math Scores


```{r, fig.height= 7, fig.width = 10}

# 3 Scatter Plots of Non-Averages VS. Averages, District Size & Math Scores

All.DS.Scatter <- ggplot(data = KMS, aes(x = dis_size, y = mathn))+
  geom_point(color = "dodgerblue")+
  stat_smooth(method = "lm", color = "gold1", se = F)+
  xlab("Centered ln() District=Level Size")+ 
  ylab("Math Score")+ 
  labs(title = "Math Score by District-Level Size All 48,058 Kentucky Middle Schoolers")

Avg.DS.Scatter <- ggplot(data = DISTRICT.KMS, aes(x = MEAN.DLS, y = MEAN.MATH))+
  geom_point(color = "magenta1")+
  stat_smooth(method = "lm", color ="springgreen2", se = F)+
  xlab("Average Centered ln() District-Level Size")+ 
  ylab("Average Math Score")+ 
  labs(title = "Average Math Score by Average District-Level Size for Middle School Students in Kentucky")

grid.arrange(All.DS.Scatter, Avg.DS.Scatter, ncol = 1)


```


### 4 Histograms of Non-Averages VS. Averages in Math Scores

```{r, fig.width=10}
# 4 Histograms of Non-Averages VS. Averages in Math Scores

All.KMS.Histogram <- ggplot(data = KMS, aes(x = mathn))+
  geom_histogram(color = "blue", fill = "pink", bins = 40)+
  xlab("Math Score")+ 
  ylab("Frequency")+ 
  labs(title = "Distribution of Math Scores for All 48,058 Middle Schoolers in Kentucky")

Avg.KMS.Histogram <- ggplot(data = DISTRICT.KMS, aes(x = MEAN.MATH))+
  geom_histogram(color = "red", fill = "orange", bins = 40)+
  xlab("Average Math Score")+ 
  ylab("Frequency")+ 
  labs(title = "Distribution of Average Math Scores by District for Middle School Students in Kentucky")

grid.arrange(All.KMS.Histogram, Avg.KMS.Histogram, ncol = 1)
```


### 5 Box Plots of Math Scores by Ethnicity and Sex for All Kentucky Middle School Students In Our Data

```{r, fig.width = 10, fig.height = 6}

# Box Plots of Math Scores by Ethnicity and Sex for All Kentucky Middle School Students In Our Data



KMS.Ethnicity.Sex <- KMS %>% mutate(female = ifelse(female == 1, "F", "M"), nonwhite = ifelse(nonwhite == 1, "NW", "W")) %>% unite(col = "Eth.Sex", "female", "nonwhite", sep = "/")

KMS.Ethnicity.Sex$Eth.Sex <- as.character(KMS.Ethnicity.Sex$Eth.Sex)

ggplot(data = KMS.Ethnicity.Sex, aes(x = Eth.Sex, y = mathn, fill = Eth.Sex))+
  geom_boxplot(size = 1, color = "gold1")+
  xlab("Sex & Ethnicity")+ 
  ylab("California Math Assessment Score")+ 
  labs(title = "Comparison of California Math Assessment Scores for the Combination of Female, Male, White, 
       and Non-White Kentucky 8th Grade Students")

```



NOTE: I omitted this second portion of the report because it is unnecessary for this assignment. 




